\documentclass[handout]{beamer}

%\let\Tiny\tiny% http://tex.stackexchange.com/a/94159/5764
\setbeamertemplate{navigation symbols}{}

\usepackage[outputdir=build]{minted}
\definecolor{almostnogray}{rgb}{0.92,0.92,0.92}
\setminted{breaklines=true,breakanywhere=true,bgcolor=almostnogray,fontsize=\scriptsize}

\title{Introduction to OpenMP}
\subtitle{Ferienakademie 2017}
\author{Nathan Brei}
\institute{}
\date\today

\begin{document}
\begin{frame}
  \titlepage
\end{frame}

\begin{frame}[fragile]
  \frametitle{OpenMP Parallel Regions}
  \begin{columns}[t]%[onlytextwidth]
    \begin{column}{0.65\textwidth}
      \inputminted{c}{src/ex1.c}
    \end{column}
    \pause
    \begin{column}{0.4\textwidth}
      \inputminted{text}{output/ex1.txt}
    \end{column}
  \end{columns}
  \pause
  \begin{itemize}
  \item \emph{Fork-join execution model}
  \item Print statements are interleaved in random order
  \end{itemize}
\end{frame}



\begin{frame}[fragile]
  \frametitle{OpenMP Hello World}
  \begin{columns}[t]%[onlytextwidth]
    \begin{column}{0.65\textwidth}
      \inputminted{c}{src/ex1.c}
    \end{column}
    \pause
    \begin{column}{0.4\textwidth}
      \begin{minted}[gobble=8]{text}
        Hello there oh weirdo
        Why is everything so strange
        Let's eat more food.
        Good night sweet prince.
      \end{minted}
    \end{column}
  \end{columns}
  \pause
  \begin{block}{Observations}
    \begin{itemize}
    \item \emph{Fork-join execution model}
    \item Print statements output in a random order
    \end{itemize}
  \end{block}
\end{frame}
\begin{frame}
  \frametitle{This is a frame title}
  \begin{columns}[onlytextwidth]
    \begin{column}{0.4\textwidth}
      \centering
      %\rule{100pt}{150pt}% Place your graphic here
      Here is some text. I hope this works. Here is some more text.
    \end{column}
    \begin{column}{0.2\textwidth}
      \centering
      \rule{40pt}{150pt}% Place your graphic here
    \end{column}
    \begin{column}{0.4\textwidth}
    Here is some regular text in a column. And there is an equation
    \[
      f(x)=ax^2+bx+c
    \]
    Here is some more text.
    \end{column}
​  \end{columns}
  \begin{block}{Takeaways}
    \begin{itemize}
    \item Nothing is as it seems, and everything is disappointing.
    \item Don't sleep too late otherwise your life ends
    \end{itemize}
  \end{block}

\end{frame}
\end{document}
